---
layout: page
title: R Fundamentals I
subtitle: Data frames and reading in data
---

```{r, include=FALSE}
source("tools/chunk-options.R")
```

> ## Learning Objectives {.objectives}
>
> * Become familiar with data frames
> * Be able to read regular data into R
>

## Data frames

Data frames are similar to matrices, except each column can be a
different atomic type. A data frames is the standard structure for
storing and manipulating rectangular data sets.  Underneath the hood,
data frames are really lists, where each element is an atomic vector,
with the added restriction that they're all the same length.  As you
will see, if we pull out one column of a data frame,we will have a
vector. You will probably find that data frames are more complicated
than vectors and other data structures we have considered, but they
provide powerful capabilities.
 

Data frames can be created manually with the `data.frame` function:

```{r}
df <- data.frame(id = c('a', 'b', 'c', 'd', 'e', 'f'), x = 1:6, y = 100:95)
df
```

As with matrices, you can use the functions `dim`, `nrow` and `ncol` to view the dimensions of the data frame.

 We can access a column
in a `data.frame` by using the `$` operator
```{r}
df$x
```
or
```{r}
df[,'x']
```

We can add columns or rows to a data.frame using `cbind` or `rbind`
(these are the two-dimensional equivalents of the `c` function):

## Adding columns to a data frame

To add a column we can use `cbind`:

```{r}
df <- cbind(df, z = 6:1, caps = LETTERS[1:6])
df
```
(`LETTERS` and `letters` are built-in constants.)

Since under the hood data frames are lists, we can add columns just like adding new elements to lists:
```{r}
df$zero <- rep(0, 6)
df
```


## Adding rows to a data frame 

To add a row we use `rbind`:

```{r}
df <- rbind(df, data.frame(id = "g", x = 11, y = 42, z = 0, caps = 'G', zero = NA))
str(df)
```


## Deleting rows and handling NA


There are multiple ways to delete a row containing `NA`:

```{r}
df[-7, ]  # The minus sign tells R to delete the row
df[!is.na(df$zero), ]  # is.na() returns TRUE where its argument is NA
subset(df, !is.na(zero)) # keep records where the expression is TRUE
df <- na.omit(df)  # Another function for the same purpose; checks all variables
df
```

## Combining data frames

We can also row-bind data.frames together:

```{r}
rbind(df, df)
```
 
## Merging data frames
Data frames can be merged on one or more columns. Create a second dataset and merge it with our `df` object:
```{r}
df2 <- data.frame(id = letters[1:5], X = 101:105)
df2
df <- merge(df, df2, by = "id", all = TRUE)
df
``` 


## Reading in data

Now we want to load our data into R. We will use PSRC land use data, particularly in this lesson the number of households in cities.
Before reading in data, it's a good idea to have a look at its structure. From a shell terminal, you can do:

```{r, engine='bash'}
cd data/ # navigate to the data directory of the project folder
head cities_households.csv
```

The file contains comma-separated values and a header row. We can use `read.table` to read this into R:

```{r}
hh <- read.table(file = "data/cities_households.csv", header = TRUE, sep = ",")
head(hh)
```

Because we know the structure of the data, we're able
to specify the appropriate arguments to `read.table`.
Without these arguments, `read.table` will do its
best to do something sensible, but it is always more
reliable to explicitly tell `read.table` the structure
of the data. `read.csv` function provides a convenient shortcut
for loading in CSV files.

```{r}
# Here is the read.csv version of the above command
hh <- read.csv(file = "data/cities_households.csv")
head(hh)
```

> ## Miscellaneous Tips {.callout}
>
> 1. Another type of file you might encounter are tab-separated
> format. To specify a tab as a separator, use `sep="\t"`.
>
> 2. You can also read in files from the Internet by replacing
> the file paths with a web address.
>
> 3. You can read directly from excel spreadsheets without
> converting them to plain text first by using the `xlsx` package.
>

To make sure our analysis is reproducible, we should put the code
into a script file so we can come back to it later. It can be then run using the `source` function, using the file path
as its argument (or by pressing the "source" button in RStudio).

## Using data frames


To recap what we've just learned, let's have a look at our example
data (number of households in various cities for various years).

Remember, there are a few functions we can use to interrogate data structures in R:

```{r, eval=FALSE}
class() # what is the data structure?
typeof() # what is its atomic type?
length() # how long is it? What about two dimensional objects?
attributes() # does it have any metadata?
str() # A full summary of the entire object
dim() # Dimensions of the object - also try nrow(), ncol()
```

Let's use them to explore the household dataset.

```{r}
class(hh)
```

The household data is stored in a "data.frame". This is the default
data structure when you read in data, and (as we've heard) is useful
for storing data with mixed types of columns.

Let's add a column with the name of each city and county and merge the two datasets:

```{r}
cities <- read.csv(file="data/cities.csv")
head(cities)
```

Let's merge the cities dataset with our households dataset.
```{r}
hh <- merge(hh, cities, by="city_id", all = TRUE)
head(hh)
```

The first thing you should do when reading data in, is check that it matches what
you expect, even if the command ran without warnings or errors. The `str` function,
short for "structure", is really useful for this:

```{r}
str(hh)
```

We can see that the object is a `data.frame` with `r nrow(hh)` observations (rows),
and `r ncol(hh)` variables (columns). Below that, we see the name of each column, followed
by a ":", followed by the type of variable in that column, along with the first
few entries.

We can retrieve or modify the column or row names
of the data.frame:

```{r}
colnames(hh) 
colnames(hh)[ncol(hh)] <- "county" # rename the last column
head(hh, n = 10)
```

```{r}
rownames(hh)[1:20]
```

See those numbers in the square brackets on the left? That tells you
the number of the first entry in that row of output. 

There are a few related ways of retrieving and modifying this information.
`attributes` will give you both the row and column names, along with the
class information, while `dimnames` will give you just the rownames and
column names.

In both cases, the output object is stored in a `list`:

```{r}
str(dimnames(hh))
```

We can look at some summary statistics:
```{r}
summary(hh)
```

