---
layout: page
title: R Fundamentals I
subtitle: Vectorisation
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/cities_households.csv", header=TRUE, sep=",")
cities <- read.csv(file="data/cities.csv")
hh <- merge(hh, cities, by="city_id", all=TRUE)
colnames(hh)[ncol(hh)] <- "county"
pierce <- subset(hh, county == "Pierce")
```

> ## Learning Objectives {.objectives}
>
> * To understand vectorised operations in R.
>

Most of R's functions are vectorised, meaning that the function will
operate on all elements of a vector without needing to loop through
and act on each element one at a time. This makes writing code more
concise, easy to read, and less error prone.


```{r}
x <- 1:4
x * 2
```

The multiplication happened to each element of the vector.

We can also add two vectors together:

```{r}
y <- 6:9
x + y
```

Each element of `x` was added to its corresponding element of `y`:

```{r, eval=FALSE}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```


Vectorised operations work element-wise on matrices:

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m * -1  
```
 

> ## Tip: element-wise vs. matrix multiplication {.callout}
>
> Very important: the operator `*` gives you element-wise multiplication!
> To do matrix multiplication, we need to use the `%*%` operator:
> 
> ```{r}
> m %*% matrix(1, nrow = 4, ncol = 1)
> matrix(1:4, nrow = 1) %*% matrix(1:4, ncol = 1)
> ```
>
> For more on matrix algebra, see the [Quick-R reference
> guide](http://www.statmethods.net/advstats/matrix.html)

To combine a matrix with a vector, keep in mind that the element-wise combination happens by columns:
```{r}
m
x
m * x
```
To do it by rows, first create a matrix from the vector and then combine the two matrices:
```{r}
xm <- matrix(rep(x, 3), nrow = 3, byrow = TRUE)
xm
m * xm
```


> ## Challenge 1 {.challenge}
>
> Given the following matrix:
>
> ```{r}
> m <- matrix(1:12, nrow = 3, ncol = 4)
> m
> ```
>
> Write down what you think will happen when you run:
>
> 1. `m ^ -1`
> 2. `m * c(1, 0, -1)`
> 3. `m > c(0, 20)`
> 4. `m * c(1, 0, -1, 2)`
>
> Did you get the output you expected? If not, ask a helper!
>



## Applying functions across rows/columns

What if we need an operation (average, sum etc.)  across rows or across columns? 

<img src="fig/r-operations-across-axes.svg" alt="Operations Across Axes" />

To support this, we can use the `apply` function.

> ## Tip {.callout}
>
> To learn about a function in R, e.g. `apply`, we can read its help
> documention by running `help(apply)` or `?apply`.

`apply` allows us to repeat a function on all of the rows (`MARGIN = 1`) or columns (`MARGIN = 2`) of a data frame. For example, using the `m` matrix, instead of
```{r}
mean(m[1,])
mean(m[2,])
mean(m[3,])
```
we do
```{r}
apply(m, 1, mean) # average across rows
```
Also, we can do
```{r}
apply(m, 2, min) # minimum across columns
apply(m, 2, sum) # sum across columns
```

> ## Tip: shortcuts for sums and means in matrices {.callout}
>
> Another way to compute row and column sums/means for matrices is
>
> ```{r, eval = FALSE}
> rowSums(m)  # equivalent to apply(m, 1, sum)
> colSums(m)  # equivalent to apply(m, 2, sum)
> rowMeans(m) # equivalent to apply(m, 1, mean)
> colMeans(m) # equivalent to apply(m, 2, mean)
> ```
> However, these shortcuts are not available for other functions like `median`, `min` or `max`. In such cases, the 
> `apply` function has to be used.
>
> The following should hold:
> ```{r}
> all(colSums(m) == apply(m, 2, sum) )
> ```


Similarly, with our `pierce` subset dataset created in the previous session, we can get household totals:
```{r}
apply(pierce[,2:6], 2, sum)
```
and the all-cities `hh` dataset:
```{r}
apply(hh[,2:6], 2, sum)
```

However, for a dataset with missing values:
```{r}
df <- data.frame(id = letters[1:5], X = c(NA, rep(100, 4)), Y = c(rep(5, 3), NA, NA))
df
```

we get:
```{r}
apply(df[,2:3], 2, sum)
```

because
```{r}
sum(c(3, NA))
```


but
```{r}
sum(c(3, NA), na.rm=TRUE)
```

Therefore
```{r}
apply(df[,2:3], 2, sum, na.rm=TRUE)
```


## Challenge solutions

> ## Solution to challenge 1 {.challenge}
>
> Given the following matrix:
>
> ```{r}
> m <- matrix(1:12, nrow = 3, ncol = 4)
> m
> ```
>
>
> Write down what you think will happen when you run:
>
> 1. `m ^ -1`
>
> ```{r, echo=FALSE}
> m ^ -1
> ```
>
> 2. `m * c(1, 0, -1)`
>
> ```{r, echo=FALSE}
> m * c(1, 0, -1)
> ```
>
> 3. `m > c(0, 20)`
>
> ```{r, echo=FALSE}
> m > c(0, 20)
> ```
>


