---
layout: page
title: R for Data Analysis
subtitle: Vectorisation
minutes: 30
---

```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/09-vectorisation-")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/city__table__households.csv", header=TRUE, sep=",")
cities <- read.table(file="data/cities.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
counties <- read.table(file="data/counties.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
hh <- merge(hh, cities, by="city_id", all=TRUE)
hh <- merge(hh, counties, by="county_id", all=TRUE)
colnames(hh)[3:6] <- c("hh10", "hh20", "hh30", "hh40")
pierce <- subset(hh, county_name == "Pierce")
```

> ## Learning Objectives {.objectives}
>
> * To understand vectorised operations in R.
>

Most of R's functions are vectorised, meaning that the function will
operate on all elements of a vector without needing to loop through
and act on each element one at a time. This makes writing code more
concise, easy to read, and less error prone.


```{r}
x <- 1:4
x * 2
```

The multiplication happened to each element of the vector.

We can also add two vectors together:

```{r}
y <- 6:9
x + y
```

Each element of `x` was added to its corresponding element of `y`:

```{r, eval=FALSE}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```


Vectorised operations work element-wise on matrices:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m * -1  
```
 

> ## Tip: element-wise vs. matrix multiplication {.callout}
>
> Very important: the operator `*` gives you element-wise multiplication!
> To do matrix multiplication, we need to use the `%*%` operator:
> 
> ```{r}
> m %*% matrix(1, nrow=4, ncol=1)
> matrix(1:4, nrow=1) %*% matrix(1:4, ncol=1)
> ```
>
> For more on matrix algebra, see the [Quick-R reference
> guide](http://www.statmethods.net/advstats/matrix.html)

To combine a matrix with a vector, keep in mind that the element-wise combination happens by columns:
```{r}
m
x
m * x
```
To do it by rows, first create a matrix from the vector and then combine the two matrices:
```{r}
xm <- matrix(rep(x, 3), nrow=3, byrow=TRUE)
xm
m * xm
```




> ## Challenge 3 {.challenge}
>
> Given the following matrix:
>
> ```{r}
> m <- matrix(1:12, nrow=3, ncol=4)
> m
> ```
>
> Write down what you think will happen when you run:
>
> 1. `m ^ -1`
> 2. `m * c(1, 0, -1)`
> 3. `m > c(0, 20)`
> 4. `m * c(1, 0, -1, 2)`
>
> Did you get the output you expected? If not, ask a helper!
>

> ## Challenge 4 {.challenge}
>
> We're interested in looking at the sum of the
> following sequence of fractions:
>
> ```{r, eval=FALSE}
>  x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
> ```
>
> This would be tedious to type out, and impossible for high values of
> n.  Use vectorisation to compute x when n=100. What is the sum when
> n=10,000?


## Applying functions across rows/columns

What if we need an operation (average, sum etc.)  across rows or across columns? 

<img src="fig/r-operations-across-axes.svg" alt="Operations Across Axes" />

To support this, we can use the `apply` function.

> ## Tip {.callout}
>
> To learn about a function in R, e.g. `apply`, we can read its help
> documention by running `help(apply)` or `?apply`.

`apply` allows us to repeat a function on all of the rows (`MARGIN = 1`) or columns (`MARGIN = 2`) of a data frame. For example, using the `m` matrix, instead of
```{r}
mean(m[1,])
mean(m[2,])
mean(m[3,])
```
we do
```{r}
apply(m, 1, mean) # average across rows
```
Also, we can do
```{r}
apply(m, 2, min) # minimum across columns
apply(m, 2, sum) # sum across columns
```

Similarly, with our pierce subset dataset:
```{r}
apply(pierce[,3:6], 2, sum)
```
and the households dataset:
```{r}
apply(hh[,3:6], 2, sum)
```
because
```{r}
sum(c(3, NA))
sum(c(3, NA), na.rm=TRUE)
```
Therefore
```{r}
apply(hh[,3:6], 2, sum, na.rm=TRUE)
```

## Challenge solutions

> ## Solution to challenge 3 {.challenge}
>
> Given the following matrix:
>
> ```{r}
> m <- matrix(1:12, nrow=3, ncol=4)
> m
> ```
>
>
> Write down what you think will happen when you run:
>
> 1. `m ^ -1`
>
> ```{r, echo=FALSE}
> m ^ -1
> ```
>
> 2. `m * c(1, 0, -1)`
>
> ```{r, echo=FALSE}
> m * c(1, 0, -1)
> ```
>
> 3. `m > c(0, 20)`
>
> ```{r, echo=FALSE}
> m > c(0, 20)
> ```
>

> ##  Challenge 4 {.challenge}
>
> We're interested in looking at the sum of the
> following sequence of fractions:
>
> ```{r, eval=FALSE}
>  x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
> ```
>
> This would be tedious to type out, and impossible for
> high values of n.
> Can you use vectorisation to compute x, when n=100?
> How about when n=10,000?
>
> ```{r}
> sum(1/(1:100)^2)
> sum(1/(1:1e04)^2)
> n <- 10000
> sum(1/(1:n)^2)
> ```
> 
> We can also obtain the same results using a function:
> ```{r}
> inverse_sum_of_squares <- function(n) {
>   sum(1/(1:n)^2)
> }
> inverse_sum_of_squares(100)
> inverse_sum_of_squares(10000)
> n <- 10000
> inverse_sum_of_squares(n)
> ```
>

