---
layout: page
title: R Fundamentals I
subtitle: Data manipulation and simple plots
---

```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/06-data-manip-")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/cities_households.csv", header=TRUE, sep=",")
cities <- read.csv(file="data/cities.csv")
hh <- merge(hh, cities, by="city_id", all=TRUE)
colnames(hh)[ncol(hh)] <- "county"
```

> ## Learning Objectives {.objectives}
>
> * To be able to summarize and visualize data.
>

## Data manipulation

Let's try vectorisation on the `hh2050` column of the `hh` dataset.
Make a new column in the `hh` data frame that
contains households in units of thousands of people.
```{r}
hh$hh50inT <- hh$hh2050 / 1000
head(hh)
```
Create a log transformation:
```{r}
hhlog <- hh
hhlog[,2:6] <- log(hhlog[,2:6])
head(hhlog)
```

Create a dataset of differences:
```{r}
hhdif <- cbind(hh[, c(1,7:9)], hh[,3:6] - hh[,2:5])
head(hhdif)
summary(hhdif)
```

## Scatter plot and histogram

Plot results:
```{r}
plot(hh$hh2016, hh$hh2050, xlab = "2016", ylab = "2050")
abline(0,1)
plot(hh$hh2016, hh$hh2050, xlab = "2016", ylab = "2050", log = "xy")
abline(0,1)
hist(hh$hh2050 - hh$hh2016, main="Change 2050-2016")
abline(v = 0, lwd = 2, col = "red")
```

Investigate records more than 200% increase from 2016 to 2050:
```{r}
lrg <- subset(hh, (hh2050 - hh2016)/hh2016 * 100 > 200)
dim(lrg)
plot(lrg$hh2016, lrg$hh2050, xlab = "2016", ylab = "2050", log = "xy")
text(lrg$hh2016, lrg$hh2050, labels = lrg$city_name, pos = 3)
abline(0,1)
```

